PGxMine: Text mining for curation of PharmGKB

### 2019-10-01 paper/paper.Rmd
Updates to paper and paperover code to export to PSB format

```{r pgxmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('figures.R')
source('pharmgkb_comparison.R')
source('subsections.R')
invisible(dev.off())
```

TODO: Authors

@@

Precision medicine has been described as getting patients "the right drug at the right dose at the right time" [@hamburg2010path]. It has been shown that genetics can play a significant part in whether the drug and dose are right for that particular patient and numerous diagnostic tests have been developed to test for genetic variants related to drug efficacy [@relling2015pharmacogenomics]. These pharmacogenomic variants encompass germline variants or somatic variants in cancer. They may affect the direct target of a drug, potentially by disrupting binding pockets, (e.g. T790M for many EGFR inhibitors [@zhou2009novel]), the metabolizing enzymes that may affect drug concentrations, by affecting enzyme efficiency (e.g. the cytochrome P450 family of enzymes [@tornio2018cytochrome]), or drug transporters that may affect the disposition of the drugs and influence efficacy and toxicity.  Response to several drugs is now known to be affected greatly by individual genetic diversity. For example, the dosing of the blood-thinner warfarin is known to be influenced by variation in CYP2C9, VKORC1, and CYP4F2 genes [@johnson2017clinical].

Cataloging this expanding knowledge of pharmacogenomic variants is the goal of the PharmGKB. Launched in 2000 [@whirl2012pharmacogenomics], it collects, curates and disseminates knowledge about gene-drug associations, many of which are clinically actionable, and provides resources for both researchers and clinicians. A group of expert curators searches the scientific literature for relevant pharmacogenomic papers and add their findings to the knowledge base. However, managing the curation of a biomedical knowledge base is a very time-consuming and challenging task. As precision medicine efforts increase and the cost of sequencing decreases, more genetic biomarkers will be found that might prove to be clinically actionable. It is already a huge challenge to identify relevant papers and these technological developments will inevitably lead to an increase in the number of published papers and an increased burden on the PharmGKB curators.

@@

To extract pharmacogenomic variants associated with specific chemicals, we propose to use a biomedical relation extraction method. The field of biomedical relation extraction focuses on extracting mentions of particular relations between entities (e.g. genes, drugs, variants, etc) in unstructured text. Dependency-parse based methods have seen great success in this area [@lever2017painless] while deep learning methods achieve similar performance but face issues with overfitting [@mehryary2016deep]. Relation extraction relies on first identifying mentions of specific entities, e.g. chemicals and variants, using entity extraction tools such as BANNER [@leaman2008banner] and tmVar [@wei2017tmvar]. The Pubtator resource has provided PubMed abstracts annotated with chemicals, diseases, genes, species, and variants [@wei2013pubtator]. Numerous projects have built on top of the annotations provided by PubTator [@percha2018global,@mahmood2016dimex]. Recently, the Pubtator Central (PTC) resource expanded this to open-access Pubmed Central full-text articles [@wei2019pubtator].

In this paper, we present PGxMine, a text-mined resource of pharmacogenomic associations between chemicals and variants. We build upon the PubTator Central resource and use a text-alignment method to extract specific mentions of chemicals, genes, and variants in the text. We further identify star alleles that are particularly important/unique and frequently appears in pharmacogenomic literature (e.g. CYP2D6\*2) in text. We parse and extract sentences that mention a chemical and variant and filter to enrich for sentences that likely discuss pharmacogenomic associations. We build classifiers to extract pharmacogenomic associations and apply these to all sentences accessible from PubMed abstracts and PubMed Central full-text papers. These results are then presented to the PharmGKB scientific curators to assist in their curation efforts.

@@

We downloaded the 21st August 2019 release of the Pubtator Central data which contains PubMed IDs associated with text mentions of chemicals, variants, genes, and other entities. It uses TaggerOne [@leaman2016taggerone] for chemicals, tmVar 2.0 [@wei2017tmvar] for variants and GNormPlus [@wei2015gnormplus] for genes. It provides mappings for chemical mentions to MeSH terms, genes to Entrez Gene IDs and where possible, variants to dbSNP IDs [@sherry2001dbsnp] (also known as rs IDs). We used the PubRunner infrastructure to download abstracts from PubMed and full-text papers from the PubMed Open Access (PMCOA) subset and PubMed Central Author Manuscript Collection (PMCAMC) and convert to BioC format [@comeau2013bioc].

To map chemicals in PubTator Central to the identifiers used in PharmGKB, we used DrugBank [@wishart2017drugbank]. We filtered the drug list by several categories in DrugBank to remove biological molecules that were not relevant drugs (details in Supplementary Materials). We also removed short terms (shorter than 4 characters). Variants were filtered using a custom set of exclusion words to remove common terms that appear to be variants but are not, e.g. for common cell-line names.

@@

```{r alignment, echo=F, eval=T, fig.width=10, fig.asp=0.4,  fig.cap='(ref:alignment)'}
knitr::include_graphics("fig_alignment.png")
```

(ref:alignment) Data from PubMed Central (PTC) is shown in the table and contains five columns for the PubMed ID, entity type, normalized ID, mention and tool used for extraction. The alignment of this example data is shown to text from the corresponding paper.

Unlike PubTator, PubTator Central does not provide the specific offsets of each biomedical entity within the text of the abstract or full paper. It provides the substring of the mention along with the PubMed ID. We used this information to align the PTC substrings to the corresponding text in the abstract or full-text paper wherever appropriate as shown in Figure \@ref(fig:alignment). The alignment algorithm takes each text mention and creates a regular expression to search the corresponding article text. Each regular expression starts and ends with word-boundaries so that matches wouldn’t happen inside words. We replaced each section of whitespace with a “\\s+” regular expression to map to any length of whitespace. PTC appears to translate the papers from Unicode to ASCII and provides ASCII mentions. Several important entities contain Unicode characters, e.g. TGF-α. so we replaced ASCII Greek letters (e.g. alpha) with a regex to map to either the ASCII text or the Greek character. These regular expressions then find candidate mappings for each text. We process the mappings largest to smallest so that no entity is mapped inside a larger entity. This tool produced the final document in the BioC format with annotations for the corresponding aligned entities with normalized IDs from PubTator Central.

@@

PubTator Central uses the tmVar 2 tool for extracting variant mentions [@wei2017tmvar]. It can extract DNA-level mutations (e.g. c.93G>A), protein-level mutations (e.g. T790M) and dbSNP IDs (e.g. rs12979860). Where possible, tmVar 2 will map the DNA/protein variants back to a dbSNP ID using co-mentions of genes in the document to filter down candidate matches. Hence a mention of EGFR alongside T790M provides enough context to map it to rs121434569. We are primarily interested in those that can be mapped back to dbSNP IDs but kept all variant mentions as long as they contained a specific coordinate (i.e. filtering out cases such as c.G>A). For variants that are successfully mapped to a dbSNP ID by tmVar 2, PTC does not provide the normalized text form of the variant. For example, ‘an EGFR threonine-to-methionine mutation at residue 790’ is mapped successfully to rs121434569 but we also need the HGVS normalized form, p.T790M. We implemented a re-normalization stage using regular expressions to capture the variant types output by tmVar 2.

@@

We extract all sentences that mention at least one variant (including star alleles) and one chemical from our filtered list. In an initial trial of this system focussing only on mentions of specific dbSNP IDs, we found that sentences were highly likely to mention pharmacogenomic events. However, when all variants were included, sentences were much less likely to discuss a pharmacogenomic association. Any machine learning system would then have a large challenge with such an imbalanced problem. We, therefore, filtered with a customized set of keywords (in Supplementary Materials) that increased the likelihood of a pharmacogenomic association or for a mention of a specific dbSNP ID.

@@

Due to a class balance difference, we split the dataset into two. The first set is all potential chemical/variant where the variant is a specific rs ID or a star allele. The second set is all remaining chemical/variant associations which contain variants for specific DNA or protein changes. We worked on the two sets separately because, empirically, the first set has a higher likelihood for discussing pharmacogenomic associations. For each set, we annotated 500 sentences for pharmacogenomic associations. Chemical/variant associations were annotated as pharmacogenomic if the variant was discussed as physically interacting with the chemical, as affecting the metabolism, concentration or resistance of the chemical, as causing adverse events related to the chemical, as related to addiction to the chemical, or as part of a clinical test for that chemical use. The PharmGKB curators were also interested in variants found to be negatively associated pharmacogenomically with a chemical. Thus, we also annotated negative discoveries as positive sentences in our data set. In total, we had 1000 annotated sentences which for evaluation are split into 80%/20% training and test sets. We include statistics on the annotated dataset in the supplementary materials.

We then created two relation classifiers, one for each for the two sets. We used a logistic regression as the classifier using Kindred’s default set of dependency-path based features. The logistic regression provides a thresholdable output score to control the precision-recall tradeoff. As in previous work, we used a high-precision/low-recall classifier [@lever2019cancermine]. We created precision-recall curves to evaluate the trade-off and select a threshold of 0.75 for both classifiers. We filtered out cancer-associated chemicals (using DrugBank categories) that are associated with a protein/DNA variant as they are likely somatic events which are outside the focus of the PharmGKB resource.

@@

We then used the PubRunner infrastructure to apply these two classifiers across all the aligned sentences [@anekalla2017pubrunner]. This enabled the use of a compute cluster to quickly classify sentences as to whether they contain pharmacogenomic information. We then outputted relations along with the normalized form of the chemical and genes with other metadata. The gene information was included for star alleles and for variants normalized to dbSNP if it could be extracted from dbSNP. Metadata from the source paper is also included such as PubMed ID, title, journal, publication year and section within the paper. We also provide a version of the sentence which has HTML characters escaped and the entities in the relation highlighted with basic HTML tags for easy viewing. This output is the unfiltered version available.

@@

We mapped the PGxMine set of associations to those found in PharmGKB. PharmGKB already contains some mappings directly to MeSH IDs but these needed to be supplemented with additional mappings from other sources. DrugBank provides a mapping from the MeSH ID used by PubTator Central to PharmGKB chemical IDs. Variants that have been mapped to dbSNPs or are star alleles can be mapped directly into PharmGKB. PharmGKB contains compound associations where multiple chemicals are associated with a variant. For simplicity, we unfold these associations so that all associations are from one chemical to one variant, allowing matching against the PGxMine data. For this comparison, we remove suballeles from star alleles (e.g. CYP3A5\*3A -> CYP3A5\*3). We also built a mapping for star alleles that map directly to a single rs ID to check if either was found in PharmGKB (e.g. POR\*28 maps to rs1057868). Further details of this mapping are explained in the Supplementary Materials. For each PGxMine association, we then check if the variant has been seen in PharmGKB, the chemical has been seen in PharmGKB and whether this particular association has been seen.

@@

```{r prcurves, echo=F, eval=T, fig.width=6, fig.asp=1,  fig.cap='(ref:prcurves)'}
grid.arrange(fig_prcurves)
```

@@

```{r subsectionPlot, echo=F, eval=T, fig.width=10, fig.asp=.5,  fig.cap='(ref:subsectionPlot)'}
grid.arrange(subsectionPlot)
```

(ref:subsectionPlot) From the full-text papers accessed through PubMed Central, associations are extracted from all sections of the full-text paper.

As would be expected, there is a long-tail of associations with `r paper.singlePapers`% of the unique associations being discussed in only a single paper. These singletons are of interest to the PharmGKB curators as the aim to improve coverage for rarer and less frequently discussed associations. We also find that the associations were found across all subsections of papers (Figure \@ref(fig:subsectionPlot)).

@@

However, the main limiting factor is the poor access to full-text papers for text mining. `r paper.mentionsInFullTextPerc`% (`r paper.mentionCount`/`r paper.mentionsInFullText`) mentions of pharmacogenomic associations were extracted from the main text of the article. Only a fraction of the biomedical literature is accessible for text mining. With only `r paper.pharmgkbPaperHasFullTextPerc`% (`r paper.pharmgkbPaperHasFullText`/`r paper.pharmgkbPaperNum`) of the papers curated in PharmGKB having full-text available, it is understandable that PGxMine fails to catch many of the previously curated papers. PGxMine is also limited to associations found within the same sentence. This is a problem that all current information extraction systems face as the false positive rate explodes for systems that attempt to extract relations across sentences [@lever2016verse]. This is a ripe area for further research and an annotated coreference dataset may push forward performance in this area in the future.

The tradeoff of recall and precision is a challenging one for many machine learning problems. Curators become frustrated with a system that has too many false positives and are understanding of a system that does not have 100% coverage of associations. We, therefore, target high-precision low-recall for this problem, especially when the frequently discussed associations are found in hundreds of papers, meaning that they will likely be discovered by our pipeline. For example, the most frequently extracted association is `r paper.topAssoc_chemical` with `r paper.topAssoc_variant` in `r paper.topAssoc_paperCount` papers. This means we should capture frequently discussed associations that aren't yet curated into PharmGKB. We find that the vast majority of extracted associations (`r paper.singlePapers`%) are only mentioned in one paper. While more frequently discussed associations are of greater interest to PharmGKB, these singletons provide a valuable trove for further curation. The most common mistake that the classifier makes is incorrectly matching a drug with a variant in a different clause of a sentence. This occurs most frequently in sentences with multiple drugs and variants.

Another challenge is capturing diplotypes of alleles or more complicated haplotype arrangements. While these are rare, they provide useful information for curators. We are also planning to investigate providing a feedback mechanism for curators so that they can flag issues with the information extraction pipeline. This will require an approach that allows the curators to provide for nuanced feedback for the machine learning system to guide and improve its relevance.

>>>>

```{r pgxmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('figures.R')
source('pharmgkb_comparison.R')
invisible(dev.off())
```

@@

Precision medicine has been described as getting patients "the right drug at the right dose at the right time" [@hamburg2010path]. It has been shown that genetics can play a significant part in whether the drug and dose are right for that particular patient and numerous diagnostic tests have been developed to test for genetic variants related to drug efficacy [@relling2015pharmacogenomics]. These pharmacogenomic variants encompass germline variants or somatic variants in cancer. They may affect the direct target of a drug, potentially by disrupting binding pockets, (e.g. T790M for many EGFR inhibitors [@zhou2009novel]), the metabolizing enzymes that may affect drug concentrations, by affecting enzyme efficiency (e.g. the cytochrome P450 family of enzymes [@tornio2018cytochrome]), or drug transporters that may affect the disposition of the drugs and influence efficacy and toxicity.

<!--Response to several drugs is now known to be affected greatly by individual genetic diversity. For example, the dosing of the blood-thinner warfarin is known to be influenced by variation in CYP2C9, VKORC1, and CYP4F2 genes [@johnson2017clinical]. -->

Cataloging this expanding knowledge of pharmacogenomic variants is the goal of the PharmGKB [@whirl2012pharmacogenomics]. Launched in 2000, it collects, curates and disseminates knowledge about gene-drug associations, many of which are clinically actionable, and provides resources for both researchers and clinicians. A group of expert curators searches the scientific literature for relevant pharmacogenomic papers and add their findings to the knowledge base. However, managing the curation of a biomedical knowledge base is a very time-consuming and challenging task. As precision medicine efforts increase and the cost of sequencing decreases, more genetic biomarkers will be found that might prove to be clinically actionable. It is already a huge challenge to identify relevant papers and these technological developments will inevitably lead to an increase in the number of published papers and an increased burden on the PharmGKB curators.

@@

To extract pharmacogenomic variants associated with specific chemicals, we propose to use a biomedical relation extraction method. The field of biomedical relation extraction focuses on extracting mentions of particular relations between entities (e.g. genes, drugs, variants, etc) in unstructured text. Dependency-parse based methods have seen great success in this area while deep learning methods achieve similar performance but face issues with overfitting [@deleger2016overview]. Relation extraction relies on first identifying mentions of specific entities, e.g. chemicals and variants, using entity extraction tools such as BANNER [@leaman2008banner] and tmVar [@wei2017tmvar]. The PubTator resource has provided PubMed abstracts annotated with chemicals, diseases, genes, species, and variants [@wei2013pubtator]. Numerous projects have built on top of the annotations provided by PubTator [@percha2018global,@mahmood2016dimex]. Recently, the Pubtator Central resource expanded this to open-access Pubmed Central full-text articles [@wei2019pubtator].

In this paper, we present PGxMine, a text-mined resource of pharmacogenomic associations between chemicals and variants. We build upon the PubTator Central resource and use a text-alignment method to extract specific mentions of chemicals, genes, and variants in the text. We further identify star alleles that are particularly important and frequently appear in pharmacogenomic literature (e.g. CYP2D6\*2) in text. We parse and extract sentences that mention a chemical and variant and filter to enrich for sentences that likely discuss pharmacogenomic associations. We build classifiers to extract pharmacogenomic associations and apply these to all sentences accessible from PubMed abstracts and PubMed Central full-text papers. These results are then presented to the PharmGKB scientific curators to assist in their curation efforts.

@@

We downloaded the 21st August 2019 release of the Pubtator Central data which uses TaggerOne [@leaman2016taggerone] for chemicals, tmVar 2.0 [@wei2017tmvar] for variants and GNormPlus [@wei2015gnormplus] for genes. It provides mappings for chemical mentions to MeSH terms, genes to Entrez Gene IDs and where possible, variants to dbSNP IDs [@sherry2001dbsnp] (also known as rs IDs). We used the PubRunner infrastructure to download abstracts from PubMed and full-text papers from the PubMed Open Access (PMCOA) subset and PubMed Central Author Manuscript Collection (PMCAMC) and convert to BioC format [@comeau2013bioc]. To map chemicals in PubTator Central to the identifiers used in PharmGKB, we used DrugBank [@wishart2017drugbank]. We filtered the drug list by several categories in DrugBank to remove biological molecules that were not relevant drugs (details in supplementary materials). We also removed short terms (shorter than 4 characters). Variants were filtered using a custom set of exclusion words to remove common terms that appear to be variants but are not, e.g. for common cell-line names.

@@

<!--
```{r alignment, echo=F, eval=T, fig.width=10, fig.asp=0.4,  fig.cap='(ref:alignment)'}
knitr::include_graphics("fig_alignment.png")
```

(ref:alignment) Data from PubTator Central is shown in the table and contains five columns for the PubMed ID, entity type, normalized ID, mention and tool used for extraction. The alignment of this example data is shown to text from the corresponding paper.
-->

Unlike PubTator, PubTator Central does not provide the specific offsets of each biomedical entity within the text of the abstract or full paper. It provides the substring of the mention along with the PubMed ID. We used this information to align the substrings to the corresponding text in the abstract or full-text paper wherever appropriate (with example shown in Figure S1). The alignment algorithm takes each text mention and creates a regular expression to search the corresponding article text. Each regular expression starts and ends with word-boundaries so that matches wouldn’t happen inside words. We replaced each section of whitespace with a “\\s+” regular expression to map to any length of whitespace. PubTator Central appears to translate the papers from Unicode to ASCII and provides ASCII mentions. Several important entities contain Unicode characters, e.g. TGF-α. so we replaced ASCII Greek letters (e.g. alpha) with a regex to map to either the ASCII text or the Greek character. These regular expressions then find candidate mappings for each text. We process the mappings largest to smallest so that no entity is mapped inside a larger entity.

@@

PubTator Central uses the tmVar 2.0 tool for extracting variant mentions [@wei2017tmvar]. It can extract DNA-level mutations (e.g. c.93G>A), protein-level mutations (e.g. T790M) and dbSNP IDs (e.g. rs12979860). Where possible, tmVar 2.0 will map the DNA/protein variants back to a dbSNP ID using co-mentions of genes in the document to filter down candidate matches. Hence a mention of PON1 alongside Q192R provides enough context to map it to rs662. We are primarily interested in those that can be mapped back to dbSNP IDs but kept all variant mentions as long as they contained a specific coordinate (i.e. filtering out cases such as c.G>A). For variants that are successfully mapped to a dbSNP ID by tmVar 2.0, PubTator Central does not provide the normalized text form of the variant. For example, ‘a PON1 glutamine-to-arginine mutation at residue 192’ is mapped successfully to rs662 but we also need the HGVS normalized form, p.Q192R. We implemented a re-normalization stage using regular expressions to capture the variant types output by tmVar 2.0.

@@

We extract all sentences that mention at least one variant (including star alleles) and one chemical from our filtered list. In an initial trial of this system focussing only on mentions of specific dbSNP IDs, we found that sentences were highly likely to mention pharmacogenomic events. However, when all variants were included, sentences were much less likely to discuss a pharmacogenomic association. Any machine learning system would then have a large challenge with such an imbalanced problem. We, therefore, filtered with a customized set of keywords (in supplementary materials) that increased the likelihood of a pharmacogenomic association or for a mention of a specific dbSNP ID.

@@

Due to a class balance difference, we split the dataset into two. The first set is all potential chemical/variant where the variant is a specific rs ID or a star allele. The second set is all remaining chemical/variant associations which contain variants for specific DNA or protein changes. We worked on the two sets separately because, empirically, the first set has a higher likelihood for discussing pharmacogenomic associations. For each set, we annotated 500 sentences for pharmacogenomic associations. Chemical/variant associations were annotated as pharmacogenomic if the variant was discussed as physically interacting with the chemical, as affecting the metabolism, concentration or resistance of the chemical, as causing adverse events related to the chemical, as related to addiction to the chemical, or as part of a clinical test for that chemical use. The PharmGKB curators were also interested in variants found to be negatively associated pharmacogenomically with a chemical. Thus, we also annotated negative discoveries as positive sentences in our data set. For instance, "MDR1 C3435T does not affect the metabolism of telmisartan" would be annotated as pharmacogenomically associated. In total, we had 1000 annotated sentences which for evaluation are split into 80%/20% training and test sets. We include statistics on the annotated dataset in the supplementary materials.

We then created two Kindred relation classifiers[@lever2017painless], one for each for the two sets. We used a logistic regression as the classifier using Kindred’s default set of dependency-path based features. The logistic regression provides a thresholdable output score to control the precision-recall tradeoff. As in previous work, we used a high-precision/low-recall classifier [@lever2019cancermine]. We created precision-recall curves to evaluate the trade-off and select a threshold of 0.75 for both classifiers. We filtered out cancer-associated chemicals (using DrugBank categories) that are associated with a protein/DNA variant as they are likely somatic events which are outside the focus of the PharmGKB resource.

@@

We then used the PubRunner infrastructure to apply these two classifiers across all the aligned sentences [@anekalla2017pubrunner]. This enabled the use of a compute cluster to quickly classify sentences as to whether they contain pharmacogenomic information. We then outputted relations along with the normalized form of the chemical and genes and other metadata. The gene information was included for star alleles and for variants normalized to dbSNP if it could be extracted from dbSNP. Metadata from the source paper is also included such as PubMed ID, title, journal, publication year and section within the paper. We also provide a version of the sentence which has HTML characters escaped and the entities in the relation highlighted with basic HTML tags for easy viewing. This output is the unfiltered version available.

@@

We mapped the PGxMine set of associations to those found in PharmGKB. PharmGKB already contains some mappings directly to MeSH IDs but these needed to be supplemented with additional mappings from other sources. DrugBank provides a mapping from the MeSH ID used by PubTator Central to PharmGKB chemical IDs. Variants that have been mapped to dbSNPs or are star alleles can be mapped directly into PharmGKB. PharmGKB contains compound associations where multiple chemicals are associated with a variant. For simplicity, we unfold these associations so that all associations are from one chemical to one variant, allowing matching against the PGxMine data. For this comparison, we remove suballeles from star alleles (e.g. CYP3A5\*3A -> CYP3A5\*3). We also built a mapping for star alleles that map directly to a single rs ID to check if either was found in PharmGKB (e.g. POR\*28 maps to rs1057868). Further details of this mapping are explained in the supplementary materials. For each PGxMine association, we then check if the variant has been seen in PharmGKB, the chemical has been seen in PharmGKB and whether this particular association has been seen.

@@

```{r prcurves, echo=F, eval=T, fig.width=11, fig.asp=0.5,  fig.cap='(ref:prcurves)'}
grid.arrange(fig_prcurves)
#grid.arrange(thresholdPlot)
#grid.arrange(prcurvesPlot)
```

@@

As would be expected, there is a long-tail of associations with `r paper.singlePapers`% of the unique associations being discussed in only a single paper. These singletons are of interest to the PharmGKB curators as the aim to improve coverage for rarer and less frequently discussed associations. We also find that the associations were found across all subsections of papers (Figure S2).

@@

The main limiting factor is the poor access to full-text papers for text mining. `r paper.mentionsInFullTextPerc`% (`r paper.mentionsInFullText`/`r paper.mentionCount`) mentions of pharmacogenomic associations were extracted from the main text of the article. Only a fraction of the biomedical literature is accessible for text mining. With only `r paper.pharmgkbPaperHasFullTextPerc`% (`r paper.pharmgkbPaperHasFullText`/`r paper.pharmgkbPaperNum`) of the papers curated in PharmGKB having full-text available, it is understandable that PGxMine fails to catch many of the previously curated papers. PGxMine is also limited to associations found within the same sentence. This is a problem that all current information extraction systems face as the false positive rate explodes for systems that attempt to extract relations across sentences [@lever2016verse].

<!-- This is a ripe area for further research and an annotated coreference dataset may push forward performance in this area in the future. -->

The tradeoff of recall and precision is a challenging one for many machine learning problems. Curators become frustrated with a system that has too many false positives and are understanding of a system that does not have 100% coverage of associations. We, therefore, target high-precision low-recall for this problem, especially when the frequently discussed associations are found in hundreds of papers, meaning that they will likely be discovered by our pipeline. For example, the most frequently extracted association is `r paper.topAssoc_chemical` with `r paper.topAssoc_variant` in `r paper.topAssoc_paperCount` papers. This means we should capture frequently discussed associations that aren't yet curated into PharmGKB. We find that the vast majority of extracted associations (`r paper.singlePapers`%) are only mentioned in one paper. While more frequently discussed associations are of greater interest to PharmGKB, these singletons provide a valuable trove for further curation. The most common mistake that the classifier makes is incorrectly matching a drug with a variant in a different clause of a sentence. This occurs most frequently in sentences with multiple drugs and variants. Another challenge is capturing diplotypes of alleles or more complicated haplotype arrangements. While these are rare, they provide useful information for curators.
